---
title: "Setup: Getting Everything Ready"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
execute:
  warning: false
  message: false
  eval: false
---

## What We're Doing

Before we can analyze any data, we need three things:

1. This workshop's files on your computer
2. R and RStudio installed and open
3. The ICE arrest and detainer data downloaded

Work through these steps now. Raise your hand if anything doesn't match what you see on the screen.

---

## Step 1: Get the Workshop Files

Go to this URL in your browser:

**github.com/[your-username]/ddp-workshop**

You'll see a page that looks like a folder of files. You do not need a GitHub account.

### If you've never used GitHub before — download the ZIP

1. Click the green **Code** button near the top right
2. Click **Download ZIP**
3. Find the downloaded file (probably in your Downloads folder) and double-click it to unzip
4. Move the unzipped folder somewhere you can find it — your Desktop or Documents is fine

### If you're comfortable with git — clone it

```bash
git clone https://github.com/[your-username]/ddp-workshop.git
```

---

## Step 2: Open the Project in RStudio

This step matters. Opening the project file — not just any R script — sets the working directory correctly so all the file paths in the lessons will work.

1. Open RStudio
2. Go to **File → Open Project**
3. Navigate to the `ddp-workshop` folder you just downloaded
4. Double-click **`ddp-workshop.Rproj`**

You'll know it worked when you see **`ddp-workshop`** in the upper-right corner of RStudio.

::: {.callout-warning}
If you just open RStudio without opening the project file, the lessons will fail when they try to load data. Always open via the `.Rproj` file.
:::

---

## Step 3: Install Packages

Packages are add-ons that give R new abilities — like reading Excel files or making charts. You only need to install them once.

In the RStudio **Console** (bottom-left pane), paste this and press Enter:

```{r}
source("setup.R")
```

Wait for it to finish. You'll see a lot of text scrolling — that's normal. When it's done you'll see:

```
All packages installed. You're ready to go.
```

---

## Step 4: Download the Data

The workshop uses real ICE enforcement records obtained through a Freedom of Information Act lawsuit.

1. Go to **deportationdata.org/data/ice.html**
2. Find the **2026-ICLI-00005** release
3. Download these six files:
   - Arrests FY24, FY25, FY26
   - Detainers FY24, FY25, FY26
4. Put all six files in the **`data/raw/`** folder inside your `ddp-workshop` folder

The files are large — around 80–200 MB each. If the download is slow, start it now and we'll move on.

---

## Step 5: Confirm Everything Works

Run this chunk. If it prints a number greater than zero, you're ready.

```{r}
library(readxl)
library(janitor)

test <- read_xlsx(
  "data/raw/2026-ICLI-00005_Arrests_FY26_20260311_Redacted.xlsx",
  skip = 6,
  n_max = 100
) |> clean_names()

cat("Rows loaded:", nrow(test), "\n")
cat("Columns:", ncol(test), "\n")
cat("\nYou're ready. Open lessons/01_r_basics_loading.qmd to begin.\n")
```

If you get an error that says `path does not exist`, the data file isn't in the right folder. Check that the `.xlsx` file is inside `data/raw/` and that the file name matches exactly.

---

## What Each Pane in RStudio Does

If RStudio is new to you, here's a quick orientation:

| Pane | Location | What it's for |
|------|----------|--------------|
| **Source** | Top-left | Where you write and read code |
| **Console** | Bottom-left | Where code runs and output appears |
| **Environment** | Top-right | Shows the data and variables you've created |
| **Files / Plots** | Bottom-right | Browse files; see charts when you make them |

During the workshop, you'll mostly work in the **Source** pane — reading through the lesson file and running chunks one at a time by pressing **Cmd+Enter** (Mac) or **Ctrl+Enter** (Windows).
