---
title: "Lesson 1: Loading and Cleaning ICE Arrest Data"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
execute:
  warning: false
  message: false
---

## Reporter Questions

Before we touch any code, here's what this lesson will help you answer:

- **How many people did ICE arrest in my state?**
- **What methods is ICE using — jail cooperation, or agents picking up people in the community?**
- **How does this year compare to last year?**

By the end of this lesson, you'll have a filtered dataset for your coverage area and a basic count of arrests by method.

---

## Overview

The Deportation Data Project publishes ICE enforcement records obtained through a Freedom of Information Act lawsuit. The data comes as Excel files — large, messy, and formatted for government systems, not journalists. This lesson walks you through loading and cleaning one, then stacking three fiscal years together for a before/after comparison.

**A note on the data:** These files contain both custodial arrests (ICE picks someone up from a jail) and non-custodial arrests (ICE picks someone up in the community). We'll learn to distinguish them in Lesson 2.

---

## Step 0: Setup

Before anything else, make sure you're working inside the project. In RStudio, you should see `ddp-workshop` in the upper-right corner. If not, go to File → Open Project and select `ddp-workshop.Rproj`.

Load the packages we need:

```{r setup}
library(tidyverse)   # filter, count, mutate, ggplot2, and more
library(readxl)      # load Excel files
library(janitor)     # clean column names
```

Set your state here. **This is the only line you need to change** to localize this analysis to your coverage area:

```{r config}
MY_STATE <- "MARYLAND"   # ALL CAPS, as it appears in the data
MY_AOR <- "Baltimore Area of Responsibility"
```

See `reference/aor_reference.md` for the full list of state names and their ICE Areas of Responsibility.

---

## Step 1: Load One Fiscal Year File

The FY26 arrests file is a good starting point. Download it from [deportationdata.org/data/ice.html](https://deportationdata.org/data/ice.html) and put it in `data/raw/`.

Open the file in Excel first. Notice that rows 1–6 are a preamble — metadata about the release, not actual data. The real column headers start on row 7. We tell R to skip those first 6 rows:

```{r load-fy26}
arrests_fy26_raw <- read_xlsx(
  "data/raw/2026-ICLI-00005_Arrests_FY26_20260311_Redacted.xlsx",
  skip = 6
)
```

The column names loaded from Excel have spaces and capital letters — hard to work with in R. `clean_names()` converts them to lowercase with underscores:

```{r clean-names}
arrests_fy26_raw <- arrests_fy26_raw |> clean_names()

# What columns do we have?
glimpse(arrests_fy26_raw)
```

Use `head()` to see the first few rows:

```{r preview}
head(arrests_fy26_raw)
```

---

## Step 2: Fix One Column-Specific Data Issue

Most columns in this dataset use real R `NA` values for missing data. But one column — `case_threat_level` — uses the literal text `"NA"` as a string instead. That means `is.na()` won't detect those values as missing, which matters if you ever filter or summarize by threat level.

You can verify this:

```{r check-na-issue}
# real NA values
sum(is.na(arrests_fy26_raw$case_threat_level))

# string "NA" values
sum(arrests_fy26_raw$case_threat_level == "NA", na.rm = TRUE)
```

Fix it by converting the string `"NA"` to a real `NA` across all text columns — it's a one-liner and costs nothing even for columns that don't need it:

```{r fix-na}
arrests_fy26_raw <- arrests_fy26_raw |>
  mutate(across(where(is.character), ~ na_if(.x, "NA")))
```

**Dates load as timestamps, not plain dates.** `readxl` correctly reads the date column, but it includes a time component (`POSIXct`) — e.g., `"2026-03-10 23:59:36 UTC"`. For a day-level analysis, strip the time with `as_date()`:

```{r fix-dates}
arrests_fy26_raw <- arrests_fy26_raw |>
  mutate(apprehension_date = as_date(apprehension_date))
```

```{r check-dates}
class(arrests_fy26_raw$apprehension_date)
range(arrests_fy26_raw$apprehension_date, na.rm = TRUE)
```

---

## Step 3: Load All Three Fiscal Years

One year of data gives you a snapshot. Three years give you a trend — and the ability to compare before and after the inauguration (January 20, 2025).

Write a small function to load and clean one file, then apply it to all three:

```{r load-function}
load_arrests <- function(path) {
  read_xlsx(path, skip = 6) |>
    clean_names() |>
    mutate(
      across(where(is.character), ~ na_if(.x, "NA")),
      apprehension_date = as_date(apprehension_date)
    )
}
```

```{r load-all}
arrests_fy24 <- load_arrests("data/raw/2026-ICLI-00005_Arrests_FY24_20260311_Redacted.xlsx")
arrests_fy25 <- load_arrests("data/raw/2026-ICLI-00005_Arrests_FY25_20260311_Redacted.xlsx")
arrests_fy26 <- load_arrests("data/raw/2026-ICLI-00005_Arrests_FY26_20260311_Redacted.xlsx")

# Stack them into one dataframe
arrests_all <- bind_rows(arrests_fy24, arrests_fy25, arrests_fy26)

cat("Total rows before deduplication:", nrow(arrests_all), "\n")
```

---

## Step 4: Remove Duplicate Rows

The FOIA data sometimes includes exact duplicate rows — same person, same date, same everything. Remove them:

```{r dedup}
arrests_all <- arrests_all |> distinct()

cat("Total rows after deduplication:", nrow(arrests_all), "\n")
```

---

## Step 5: Filter to Your Coverage Area

National ICE data covers hundreds of thousands of arrests. We only want the ones in our coverage area.

The primary filter uses the `state` field. But some records have a blank `state` and are instead assigned to an ICE Area of Responsibility (AOR). Without an AOR fallback, you'll miss those records entirely.

The complete filter uses both:

```{r filter-state}
arrests_local <- arrests_all |>
  filter(
    state == MY_STATE |
    (is.na(state) & toa_current_duty_aor == MY_AOR)
  )

cat("Arrests in", MY_STATE, ":", nrow(arrests_local), "\n")
```

**When NOT to use the AOR fallback:** Some AORs cover multiple states. The Newark AOR, for example, covers both New Jersey and Delaware — if you're analyzing New Jersey and filter by AOR, you'll pull Delaware records too. In that case, filter by `state` only and skip the AOR line.

Check `reference/aor_reference.md` to see which states your AOR covers before deciding.

---

## Step 6: Your First Question — What Methods Are Being Used?

`count()` is one of the most useful functions in R. It counts how many times each value appears in a column:

```{r count-method}
arrests_local |>
  count(apprehension_method, sort = TRUE)
```

You should see a list of arrest methods ranked from most to least common.

---

## Your Turn

Change `MY_STATE` at the top of this file to your coverage area and re-run the notebook.

1. How many total arrests are in your area across FY24–FY26?
2. What is the most common arrest method?
3. How many rows were removed by the deduplication step?

---

## Key Takeaways

- DDP Excel files have a 6-row preamble — always use `skip = 6`
- `clean_names()` makes column names usable in R
- `bind_rows()` stacks multiple files into one dataframe — essential for multi-year analysis
- `distinct()` removes exact duplicate rows
- `filter(state == MY_STATE)` localizes the national data to your coverage area
- `count(column)` answers "how many of each thing?" — your most-used function

**Key column names to remember:**

| Column | What it contains |
|--------|-----------------|
| `apprehension_date` | Date of arrest |
| `apprehension_method` | How the arrest was made |
| `apprehension_criminality` | Criminal history at time of arrest |
| `apprehension_site_landmark` | Facility or location |
| `anonymized_identifier` | Anonymous person ID (links to detainers) |
| `state` | State (ALL CAPS) |
| `toa_current_duty_aor` | ICE Area of Responsibility |

**In Lesson 2,** we'll summarize these arrests by method and criminality, and join them to the detainer data to find out which local agencies are cooperating with ICE.
