---
title: "Lesson 2: Summarizing Arrests and Joining Detainers"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
execute:
  warning: false
  message: false
---

## Reporter Questions

- **Are most arrests in my area happening at jails, or are agents going out into the community?**
- **What share of people arrested in my area had no criminal conviction?**
- **Is my local sheriff or police department cooperating with ICE?**
- **When ICE asks my county jail to hold someone, does the jail say yes?**

---

## Overview

In Lesson 1, you loaded the arrests data and filtered it to your state. Now we'll summarize it — finding the top facilities, breaking down arrests by type, and connecting the arrests to detainer records to see which local agencies are cooperating with ICE.

---

## Setup

```{r setup}
library(tidyverse)
library(readxl)
library(janitor)

MY_STATE <- "MARYLAND"   # change to your state
```

Re-run the loading code from Lesson 1 to get `arrests_local`. (In a real workflow, you'd save this to a file and load it here. For the workshop, re-running is fine.)

```{r load-data}
load_arrests <- function(path) {
  read_xlsx(path, skip = 6) |>
    clean_names() |>
    mutate(
      across(where(is.character), ~ na_if(.x, "NA")),
      apprehension_date = as_date(apprehension_date)
    )
}

arrests_local <- bind_rows(
  load_arrests("data/raw/2026-ICLI-00005_Arrests_FY24_20260311_Redacted.xlsx"),
  load_arrests("data/raw/2026-ICLI-00005_Arrests_FY25_20260311_Redacted.xlsx"),
  load_arrests("data/raw/2026-ICLI-00005_Arrests_FY26_20260311_Redacted.xlsx")
) |>
  distinct() |>
  filter(state == MY_STATE)
```

---

## Part A: Summarizing Arrests

### Custodial vs. Non-Custodial

The `apprehension_method` column tells us how each arrest was made. Let's start by counting every value:

```{r count-method}
arrests_local |>
  count(apprehension_method, sort = TRUE)
```

There are many values. For most stories, we want to collapse them into three buckets:

- **Custodial** — ICE picks up someone already held at a jail or prison
- **Non-Custodial / At-Large** — ICE arrests someone in the community (home, street, workplace)
- **Border / Entry** — arrests at ports of entry or border crossings

```{r recode-type}
CUSTODIAL_METHODS <- c(
  "Custodial Arrest",
  "CAP Local Incarceration",
  "CAP Federal Incarceration",
  "CAP State Incarceration",
  "287(g) Program",
  "Probation and Parole",
  "Law Enforcement Agency Response Unit",
  "Other Agency (turned over to INS)",
  "Organized Crime Drug Enforcement Task Force",
  "Criminal Alien Program"
)

AT_LARGE_METHODS <- c(
  "Non-Custodial Arrest",
  "Located",
  "ERO Reprocessed Arrest",
  "Other efforts",
  "Other Task Force",
  "Worksite Enforcement",
  "Patrol Interior"
)

arrests_local <- arrests_local |>
  mutate(arrest_type = case_when(
    apprehension_method %in% CUSTODIAL_METHODS ~ "Custodial",
    apprehension_method %in% AT_LARGE_METHODS  ~ "Non-Custodial / At-Large",
    TRUE                                        ~ "Border / Entry"
  ))
```

Now count by the three buckets:

```{r count-type}
arrests_local |>
  count(arrest_type, sort = TRUE) |>
  mutate(pct = round(n / sum(n) * 100, 1))
```

> **Caveat:** After July 2025, ICE folded CAP program arrests into `"Custodial Arrest"`, so `CAP Local Incarceration` and similar categories are mostly absent in FY26 data. The custodial total is still accurate; the breakdown between CAP and 287(g) is only reliable before July 2025. See `reference/caveats.md`.

---

### Criminality Breakdown

The `apprehension_criminality` column records whether the person had a prior conviction. Three values:

```{r count-criminality}
arrests_local |>
  count(apprehension_criminality, sort = TRUE) |>
  mutate(pct = round(n / sum(n) * 100, 1))
```

Rename the coded values to plain English:

```{r recode-criminality}
arrests_local <- arrests_local |>
  mutate(criminality_label = case_when(
    apprehension_criminality == "1 Convicted Criminal"       ~ "Convicted",
    apprehension_criminality == "2 Pending Criminal Charges" ~ "Pending Charges",
    apprehension_criminality == "3 Other Immigration Violator" ~ "No Criminal Record",
    TRUE ~ "Unknown"
  ))

arrests_local |>
  count(criminality_label, sort = TRUE) |>
  mutate(pct = round(n / sum(n) * 100, 1))
```

---

### Top Facilities

Which jails and locations are ICE arresting the most people from?

```{r top-facilities}
arrests_local |>
  filter(!is.na(apprehension_site_landmark)) |>
  count(apprehension_site_landmark, sort = TRUE) |>
  slice_head(n = 10)
```

> Some records use area codes like `"WCD GENERAL AREA"` or `"NON-SPECIFIC"` instead of a real facility name. These are real arrests but can't be mapped to a specific location. The `my_newsroom/` template excludes them from facility tables but counts them in state totals.

---

## Your Turn (Part A)

1. What percentage of arrests in your state were non-custodial (at-large)?
2. What share had no criminal conviction?
3. What are the top 10 arrest locations?

---

## Part B: Joining Detainers

A detainer is a request ICE sends to a local jail asking them to hold someone for pickup. When a jail refuses, the `detainer_lift_reason` field is recorded as `"Detainer Declined by LEA"`.

### Load the Detainers File

```{r load-detainers}
load_detainers <- function(path) {
  read_xlsx(path, skip = 6) |>
    clean_names() |>
    mutate(
      across(where(is.character), ~ na_if(.x, "NA")),
      detainer_prepare_date = as_date(detainer_prepare_date)
    )
}

detainers_all <- bind_rows(
  load_detainers("data/raw/2026-ICLI-00005_Detainers_FY24_20260311_Redacted.xlsx"),
  load_detainers("data/raw/2026-ICLI-00005_Detainers_FY25_20260311_Redacted.xlsx"),
  load_detainers("data/raw/2026-ICLI-00005_Detainers_FY26_20260311_Redacted.xlsx")
) |>
  distinct()
```

Filter detainers to your state:

```{r filter-detainers}
detainers_local <- detainers_all |>
  filter(facility_state == MY_STATE)

cat("Detainers in", MY_STATE, ":", nrow(detainers_local), "\n")
```

### What's in the Detainer Lift Reason Field?

```{r count-lift-reason}
detainers_local |>
  count(detainer_lift_reason, sort = TRUE)
```

`NA` means the detainer is still active (no outcome yet). `"Booked into Detention"` means ICE successfully picked up the person. `"Detainer Declined by LEA"` means the jail refused.

### Which Agencies Are Cooperating — and Which Are Refusing?

```{r decline-rate}
agency_summary <- detainers_local |>
  group_by(detainer_facility) |>
  summarize(
    total_detainers = n(),
    declined = sum(detainer_lift_reason == "Detainer Declined by LEA", na.rm = TRUE),
    decline_rate = round(declined / total_detainers * 100, 1)
  ) |>
  arrange(desc(total_detainers)) |>
  slice_head(n = 15)

agency_summary
```

### Join Arrests to Detainers

The `anonymized_identifier` column is an anonymous person ID that appears in both the arrests and detainers files. We can use it to link records:

```{r join-arrests-detainers}
arrests_with_detainers <- arrests_local |>
  left_join(
    detainers_local |>
      select(anonymized_identifier, detainer_prepare_date,
             detainer_facility, detainer_lift_reason),
    by = "anonymized_identifier"
  )

cat("Rows after join:", nrow(arrests_with_detainers), "\n")
```

> **Why more rows than arrests?** A `left_join` creates one row for every matching combination. If a person has multiple detainers, they'll appear multiple times. In Lesson 2 we keep it simple. The `my_newsroom/` template uses a 180-day proximity window to keep only the most relevant detainer per arrest.

---

## Your Turn (Part B)

1. What is the overall detainer decline rate in your state?
2. Which facility had the most detainers? What was its decline rate?
3. How many arrests in your state were linked to a detainer in the detainers file?

---

## Key Takeaways

- `count()` + `mutate(pct = n / sum(n) * 100)` — count and calculate percentages in one step
- `group_by()` + `summarize()` — aggregate by category (e.g., per facility)
- `slice_head(n = 10)` — keep only the top N rows
- `case_when()` — recode values; always end with `TRUE ~ "other"` as a catch-all
- `left_join()` — link two tables on a shared column; keeps all rows from the left table

**In Lesson 3,** we'll turn these numbers into charts and identify one story to pitch.
