---
title: "Lesson 3: Visualization and Story Hunting"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
execute:
  warning: false
  message: false
---

## Reporter Questions

- **Has ICE enforcement in my area spiked since the inauguration — and by how much?**
- **What does the surge look like visually, month by month?**
- **Are arrests in my area driven more by jail cooperation or by agents going into the community?**
- **What's the one number that makes my local story worth publishing?**

---

## Overview

Numbers tell you something. Charts let you see it — and let your editor see it. This lesson covers the basics of `ggplot2`, the R charting library, and ends with a story-hunting exercise using real published stories as models.

---

## Setup

```{r setup}
library(tidyverse)
library(readxl)
library(janitor)
library(scales)      # for comma() and percent() in axis labels

MY_STATE  <- "MARYLAND"
MY_LABEL  <- "Maryland"     # used in chart titles
```

Re-run the loading and cleaning code:

```{r load-data}
load_arrests <- function(path) {
  read_xlsx(path, skip = 6) |>
    clean_names() |>
    mutate(
      across(where(is.character), ~ na_if(.x, "NA")),
      apprehension_date = as_date(apprehension_date)
    )
}

CUSTODIAL_METHODS <- c(
  "Custodial Arrest", "CAP Local Incarceration", "CAP Federal Incarceration",
  "CAP State Incarceration", "287(g) Program", "Probation and Parole",
  "Law Enforcement Agency Response Unit", "Other Agency (turned over to INS)",
  "Organized Crime Drug Enforcement Task Force", "Criminal Alien Program"
)

AT_LARGE_METHODS <- c(
  "Non-Custodial Arrest", "Located", "ERO Reprocessed Arrest",
  "Other efforts", "Other Task Force", "Worksite Enforcement", "Patrol Interior"
)

arrests_local <- bind_rows(
  load_arrests("data/raw/2026-ICLI-00005_Arrests_FY24_20260311_Redacted.xlsx"),
  load_arrests("data/raw/2026-ICLI-00005_Arrests_FY25_20260311_Redacted.xlsx"),
  load_arrests("data/raw/2026-ICLI-00005_Arrests_FY26_20260311_Redacted.xlsx")
) |>
  distinct() |>
  filter(state == MY_STATE) |>
  mutate(
    arrest_type = case_when(
      apprehension_method %in% CUSTODIAL_METHODS ~ "Custodial",
      apprehension_method %in% AT_LARGE_METHODS  ~ "Non-Custodial / At-Large",
      TRUE                                        ~ "Border / Entry"
    ),
    criminality_label = case_when(
      apprehension_criminality == "1 Convicted Criminal"         ~ "Convicted",
      apprehension_criminality == "2 Pending Criminal Charges"   ~ "Pending Charges",
      apprehension_criminality == "3 Other Immigration Violator" ~ "No Criminal Record",
      TRUE ~ "Unknown"
    )
  )
```

---

## Part A: Visualization

### How ggplot2 Works

Every `ggplot2` chart is built in layers:

1. `ggplot(data, aes(x, y))` — what data and what goes on each axis
2. `geom_*()` — what shape to draw (bars, lines, points)
3. `labs()` — title, subtitle, axis labels, caption
4. `theme_*()` — overall look

Let's build a chart step by step.

### Chart 1: Monthly Arrest Trend

First, prepare the data — count arrests per month:

```{r monthly-data}
monthly_arrests <- arrests_local |>
  mutate(month = floor_date(apprehension_date, "month")) |>
  count(month)

head(monthly_arrests)
```

Now build the chart:

```{r monthly-chart}
ggplot(monthly_arrests, aes(x = month, y = n)) +
  geom_line(linewidth = 1, color = "#2c6fad") +
  geom_point(size = 2, color = "#2c6fad") +
  labs(
    title = paste("ICE Arrests in", MY_LABEL, "by Month"),
    subtitle = "FY2024–FY2026",
    x = NULL,
    y = "Arrests",
    caption = "Source: DHS via Deportation Data Project (FOIA)"
  ) +
  scale_y_continuous(labels = comma) +
  theme_minimal()
```

### Add the Inauguration Line

Every chart in this analysis should mark January 20, 2025 — the day that enforcement priorities changed:

```{r monthly-chart-inauguration}
ggplot(monthly_arrests, aes(x = month, y = n)) +
  geom_line(linewidth = 1, color = "#2c6fad") +
  geom_point(size = 2, color = "#2c6fad") +
  geom_vline(
    xintercept = as.Date("2025-01-20"),
    linetype = "dashed",
    color = "gray40"
  ) +
  annotate(
    "text",
    x = as.Date("2025-01-20"),
    y = max(monthly_arrests$n) * 0.95,
    label = "Inauguration\nJan 20, 2025",
    hjust = -0.05,
    size = 3,
    color = "gray40"
  ) +
  labs(
    title = paste("ICE Arrests in", MY_LABEL, "by Month"),
    subtitle = "FY2024–FY2026",
    x = NULL,
    y = "Arrests",
    caption = "Source: DHS via Deportation Data Project (FOIA)"
  ) +
  scale_y_continuous(labels = comma) +
  theme_minimal()
```

### Save the Chart

```{r save-chart, eval=FALSE}
ggsave("my_monthly_chart.png", width = 9, height = 5, dpi = 300)
```

### Chart 2: Custodial vs. Non-Custodial Over Time

A stacked bar chart shows how the *composition* of arrests has shifted:

```{r stacked-bar}
monthly_by_type <- arrests_local |>
  mutate(month = floor_date(apprehension_date, "month")) |>
  count(month, arrest_type)

ggplot(monthly_by_type, aes(x = month, y = n, fill = arrest_type)) +
  geom_col() +
  geom_vline(
    xintercept = as.Date("2025-01-20"),
    linetype = "dashed",
    color = "white"
  ) +
  scale_fill_manual(values = c(
    "Custodial"              = "#2c6fad",
    "Non-Custodial / At-Large" = "#c94040",
    "Border / Entry"         = "#888888"
  )) +
  labs(
    title = paste("ICE Arrests in", MY_LABEL, "by Type"),
    subtitle = "Dashed line = inauguration (Jan 20, 2025)",
    x = NULL,
    y = "Arrests",
    fill = "Arrest Type",
    caption = "Source: DHS via Deportation Data Project (FOIA)"
  ) +
  scale_y_continuous(labels = comma) +
  theme_minimal()
```

### Chart 3: Criminality Breakdown

```{r criminality-bar}
criminality_summary <- arrests_local |>
  count(criminality_label) |>
  mutate(
    pct = n / sum(n),
    criminality_label = fct_reorder(criminality_label, n)
  )

ggplot(criminality_summary, aes(x = criminality_label, y = pct)) +
  geom_col(fill = "#2c6fad") +
  geom_text(
    aes(label = percent(pct, accuracy = 1)),
    hjust = -0.1,
    size = 3.5
  ) +
  coord_flip() +
  scale_y_continuous(labels = percent, limits = c(0, 0.7)) +
  labs(
    title = paste("Criminality of ICE Arrests in", MY_LABEL),
    subtitle = "FY2024–FY2026",
    x = NULL,
    y = "Share of all arrests",
    caption = "Source: DHS via Deportation Data Project (FOIA)"
  ) +
  theme_minimal()
```

---

## Your Turn (Part A)

1. Build the monthly trend chart for your state.
2. Does it show a clear increase after January 20, 2025?
3. What share of arrests were non-custodial? Build a chart showing the trend.

---

## Part B: Story Hunting

Each of these angles has been published. Find the number in your data that makes a story for your coverage area.

---

### Story Angle 1: The Surge

**Published:** Capital News Service: "Maryland's surprising role in implementing Trump's immigration agenda" (Nov. 13, 2025)
**The number:** ICE arrested more than 2,000 people in Maryland communities in under a year — and 70% had no prior conviction.

**Code starter:**

```{r story-1}
# Total arrests before and after inauguration
inauguration <- as.Date("2025-01-20")

arrests_local |>
  mutate(era = if_else(apprehension_date >= inauguration, "Post-inauguration", "Pre-inauguration")) |>
  count(era)
```

---

### Story Angle 2: Who Is Being Arrested

**Published:** Milwaukee Journal Sentinel: "ICE arrests of people without convictions grows sixfold in Wisconsin" (Apr. 16, 2026)
**The number:** The share of non-criminal arrests spiked after January 2025.

**Code starter:**

```{r story-2}
arrests_local |>
  mutate(era = if_else(apprehension_date >= inauguration, "Post-inauguration", "Pre-inauguration")) |>
  count(era, criminality_label) |>
  group_by(era) |>
  mutate(pct = round(n / sum(n) * 100, 1))
```

---

### Story Angle 3: Local Police Cooperation

**Published:** Capital News Service: "Ban on immigration partnership would make Maryland a target for ICE street arrests, Frederick County Sheriff says" (Dec. 19, 2025)
**The number:** 119 people transferred from Maryland sheriffs' jails to ICE — a higher rate than most states.

**Code starter:**

```{r story-3, eval=FALSE}
# You need the detainers data loaded — see Lesson 2
# Which facilities declined the most?
detainers_local |>
  group_by(detainer_facility) |>
  summarize(
    total   = n(),
    declined = sum(detainer_lift_reason == "Detainer Declined by LEA", na.rm = TRUE),
    decline_rate = round(declined / total * 100, 1)
  ) |>
  arrange(desc(total)) |>
  slice_head(n = 10)
```

---

### Story Angle 4: Method Shift — Jail vs. Community

**Published:** Block Club Chicago: ICE's Illinois arrests during Trump's crackdown (Dec. 2025)

**Code starter:**

```{r story-4}
arrests_local |>
  mutate(month = floor_date(apprehension_date, "month")) |>
  count(month, arrest_type) |>
  group_by(month) |>
  mutate(pct = n / sum(n)) |>
  filter(arrest_type == "Non-Custodial / At-Large") |>
  ggplot(aes(x = month, y = pct)) +
  geom_line(color = "#c94040", linewidth = 1) +
  geom_vline(xintercept = as.Date("2025-01-20"), linetype = "dashed", color = "gray40") +
  scale_y_continuous(labels = percent) +
  labs(
    title = paste("Share of Non-Custodial Arrests in", MY_LABEL),
    x = NULL, y = "% of monthly arrests",
    caption = "Source: DHS via Deportation Data Project (FOIA)"
  ) +
  theme_minimal()
```

---

### Story Angle 5: Your Story

Pick one angle. Run the code for your state. Write one sentence in this format:

> ICE arrested **[NUMBER]** people in **[PLACE]** in **[TIME PERIOD]**, and **[KEY FINDING]** — compared to **[COMPARISON]** the year before.

---

## Key Takeaways

- `ggplot(data, aes(x, y)) + geom_col()` — the basic bar chart
- `geom_vline(xintercept = as.Date("2025-01-20"))` — add the inauguration line to any chart
- `annotate("text", ...)` — add a label to a reference line
- `floor_date(date, "month")` — round a date to the first of the month (for monthly totals)
- `scale_y_continuous(labels = comma)` — readable axis numbers
- `ggsave("chart.png", width = 9, height = 5)` — save for publication

**Take home:** The `my_newsroom/` folder has a complete template — change `MY_STATE` and `MY_AOR` in `config.R`, then knit `analyze.qmd` for a full analysis of your area.
