---
title: "Bonus Lesson: Where Do People End Up? (Detentions Data)"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
execute:
  warning: false
  message: false
---

## Reporter Questions

- **Where do people arrested in my area get sent after ICE takes them?**
- **How long are they held before being released or deported?**
- **Are detainees from my area being transferred to facilities far from their families and lawyers?**

---

## Overview

The arrests and detainers data tells you *who* ICE arrested and *how*. The detentions data tells you *what happened next* — which facilities they were sent to, how long they were held, and how they were released.

**Published example:** Capital News Service / The Daily Record: "MD immigrants detained by ICE find themselves far from home" (Dec. 31, 2025). Finding: Three out of four people detained in Maryland were eventually transferred to facilities in Louisiana or Texas.

---

## Setup

The detentions data comes as a large parquet file from the DDP. Download it from [deportationdata.org/data/ice.html](https://deportationdata.org/data/ice.html) and put it in `data/raw/`.

```{r setup}
library(tidyverse)
library(nanoparquet)   # install.packages("nanoparquet") if needed
library(scales)

MY_STATE <- "MARYLAND"
MY_AOR   <- "Baltimore Area of Responsibility"
MY_LABEL <- "Maryland"
```

---

## Load the Detentions File

Unlike the arrests files, the detentions data from DDP is already processed into a parquet file (a compressed data format that's faster than Excel for large files):

```{r load-detentions, eval=FALSE}
detentions <- read_parquet("data/raw/detention-stays-latest.parquet")
glimpse(detentions)
```

---

## Filter to Your Area

The detentions data uses `book_in_aor` (the AOR where the person was first booked) to identify local cases:

```{r filter-area, eval=FALSE}
detentions_local <- detentions |>
  filter(book_in_aor == MY_AOR)

cat("Detentions from", MY_AOR, ":", nrow(detentions_local), "\n")
```

---

## How Long Are People Held?

Calculate detention length in days:

```{r detention-length, eval=FALSE}
detentions_local <- detentions_local |>
  mutate(
    days_held = as.numeric(difftime(
      stay_book_out_date_time,
      stay_book_in_date_time,
      units = "days"
    ))
  )

summary(detentions_local$days_held)
```

---

## Where Do People Get Sent?

The `detention_facility_last` column shows where the person was held at the end of their stay (or is currently held). Compare that to the `book_in_aor` to find out-of-state transfers:

```{r transfer-analysis, eval=FALSE}
# What states are people ending up in?
detentions_local |>
  count(state_of_detention_facility_last, sort = TRUE) |>
  slice_head(n = 10)
```

---

## Transfer Rate

How many people ended up outside their home AOR?

```{r transfer-rate, eval=FALSE}
detentions_local |>
  mutate(
    transferred = detention_release_reason == "Transfer" |
                  (book_in_aor != state_of_detention_facility_last)
  ) |>
  summarize(
    total      = n(),
    transferred = sum(transferred, na.rm = TRUE),
    pct        = round(transferred / total * 100, 1)
  )
```

---

## Key Takeaways

- The detentions data uses `book_in_aor` (not `state`) to identify local cases
- `difftime()` calculates the number of days between two dates
- `detention_release_reason == "Transfer"` identifies formal transfers
- Three out of four Maryland detainees were transferred to Louisiana or Texas in 2025

**For per-capita comparison:** If you want to compare transfer rates between states, use the same denominator (detentions per population or detentions per arrest) and note that the `book_in_aor` spans multiple states for some AORs.
