---
title: "Bonus Lesson: Where Do People End Up? (Detentions Data)"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    theme: cosmo
execute:
  warning: false
  message: false
---

## Reporter Questions

- **Where do people arrested in my area get sent after ICE takes them?**
- **How long are they held before being released or deported?**
- **Are detainees from my area being transferred to facilities far from their families and lawyers?**

---

## Overview

The arrests and detainers data tells you *who* ICE arrested and *how*. The detentions data tells you *what happened next* — which facilities they were sent to, how long they were held, and how they were released.

**Published example:** Capital News Service / The Daily Record: "MD immigrants detained by ICE find themselves far from home" (Dec. 31, 2025). Finding: Three out of four people detained in Maryland were eventually transferred to facilities in Louisiana or Texas.

---

## Setup

The detentions data comes as a large parquet file from the DDP. Download it from [deportationdata.org/data/ice.html](https://deportationdata.org/data/ice.html) and put it in `data/processed/` (it's already a cleaned, derived file, not a raw FOIA release).

```{r setup}
library(tidyverse)
library(nanoparquet)   # install.packages("nanoparquet") if needed
library(scales)

MY_STATE <- "MARYLAND"
MY_AOR   <- "Baltimore Area of Responsibility"
MY_LABEL <- "Maryland"
```

---

## Load the Detentions File

Unlike the arrests files, the detentions data from DDP is already processed into a parquet file (a compressed data format that's faster than Excel for large files):

```{r load-detentions, eval=FALSE}
detentions <- read_parquet("data/processed/detention-stays-latest.parquet")
glimpse(detentions)
```

---

## Filter to Your Area

`book_in_aor` is the AOR where a person was booked into detention. But there's a catch that matters for the transfer story: this file is one row per **stay**, and a stay ends when someone changes status or facility. When ICE books a person in Baltimore and later moves them to Louisiana, that's recorded as **two separate stays** with two different `book_in_aor` values — so filtering to a single AOR throws away the very transfers you're trying to find.

To follow people, first get everyone who was ever booked in your area, then pull *all* of their stays:

```{r filter-area, eval=FALSE}
# Everyone booked into detention in your area at any point
local_ids <- detentions |>
  filter(book_in_aor == MY_AOR) |>
  distinct(unique_identifier) |>
  pull(unique_identifier)

# Every stay belonging to those people, so we can see where they went next
detentions_local <- detentions |>
  filter(unique_identifier %in% local_ids)

cat(MY_LABEL, "detainees:", length(local_ids),
    "people across", nrow(detentions_local), "stays\n")
```

---

## How Long Are People Held?

Each stay has a book-in and book-out timestamp. Subtract them for the length in days (a `NA` book-out means the person is still in custody):

```{r detention-length, eval=FALSE}
detentions_local <- detentions_local |>
  mutate(
    days_held = as.numeric(difftime(
      stay_book_out_date_time,
      stay_book_in_date_time,
      units = "days"
    ))
  )

summary(detentions_local$days_held)
```

---

## Where Do People Get Sent?

There is no state column in this file. The end-of-stay field is `detention_facility_last`, a facility **name**. For people booked in your area, that name is almost always a local hold room — the out-of-state moves live in their *later* stays. So the cleanest "where did they go" signal is the `book_in_aor` of those later stays:

```{r where-sent, eval=FALSE}
detentions_local |>
  filter(book_in_aor != MY_AOR) |>
  count(book_in_aor, sort = TRUE) |>
  slice_head(n = 10)
```

For Maryland this surfaces New Orleans (Louisiana) far ahead of everywhere else, followed by Harlingen, Houston, San Antonio, and El Paso (Texas) — the "far from home" pattern. To see specific facilities instead of regions, swap in `count(detention_facility_last, sort = TRUE)`.

---

## Transfer Rate

A person counts as transferred out of the area if any of their stays was booked in a *different* AOR than yours. Collapse to one row per person first, then summarize:

```{r transfer-rate, eval=FALSE}
detentions_local |>
  group_by(unique_identifier) |>
  summarize(
    left_area = any(book_in_aor != MY_AOR, na.rm = TRUE),
    .groups = "drop"
  ) |>
  summarize(
    people      = n(),
    transferred = sum(left_area),
    pct         = round(transferred / people * 100, 1)
  )
```

This is a conservative floor: it only catches moves that crossed into another AOR's books. Same-AOR facility changes and people still in their first stay don't count, so the true "moved at least once" share is higher.

---

## Key Takeaways

- This file is one row per **stay**, not per person; track people with `unique_identifier` or you'll miss every cross-AOR transfer
- `book_in_aor` identifies the booking area; there is no state field, so out-of-area moves show up as later stays in a different AOR
- `difftime()` calculates the number of days between book-in and book-out
- `detention_release_reason == "Transferred"` flags formal transfers (note the exact spelling)
- Following Maryland detainees this way reproduces the published finding: New Orleans and Texas AORs dominate the destinations

**For per-capita comparison:** If you want to compare transfer rates between states, use the same denominator (detentions per population or detentions per arrest) and note that an AOR can span multiple states.
