---
title: "ICE Arrests and Detainers — Local Analysis"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(readxl)
library(janitor)
library(scales)
library(here)        # resolves paths from the project root, however you run this

# here() anchors paths to the project root, so this works whether you knit or run chunks interactively
source(here("my_newsroom", "config.R"))
source(here("my_newsroom", "excluded_landmarks.R"))
```

---

## Load and Clean Data

```{r load-arrests}
load_arrests <- function(path) {
  read_xlsx(here(path), skip = 6) |>
    clean_names() |>
    mutate(
      across(where(is.character), ~ na_if(.x, "NA")),
      apprehension_date = as_date(apprehension_date)
    )
}

arrests_all <- bind_rows(
  load_arrests(ARRESTS_FY24),
  load_arrests(ARRESTS_FY25),
  load_arrests(ARRESTS_FY26)
)
n_raw_rows <- nrow(arrests_all)          # before de-duplication, for the Data Check
arrests_all <- distinct(arrests_all)
```

```{r filter-state}
# Filter modes: state_only (safest for multi-state AORs), state_or_aor (state plus blank-state AOR matches), state_and_aor (one AOR within a state, for sub-state breakouts like NYC)
arrests_local <- arrests_all |>
  filter(
    if (MY_FILTER_MODE == "state_only") {
      state == MY_STATE
    } else if (MY_FILTER_MODE == "state_and_aor") {
      state == MY_STATE & toa_current_duty_aor == MY_AOR
    } else {
      state == MY_STATE | (is.na(state) & toa_current_duty_aor == MY_AOR)
    }
  ) |>
  filter(apprehension_date <= END_DATE)   # full FY24-FY26 range; sliced into administration years below

cat(MY_LABEL, "arrests:", nrow(arrests_local), "\n")
```

```{r recode}
CUSTODIAL_METHODS <- c(
  "Custodial Arrest", "CAP Local Incarceration", "CAP Federal Incarceration",
  "CAP State Incarceration", "287(g) Program", "Probation and Parole",
  "Law Enforcement Agency Response Unit", "Other Agency (turned over to INS)",
  "Organized Crime Drug Enforcement Task Force", "Criminal Alien Program"
)

AT_LARGE_METHODS <- c(
  "Non-Custodial Arrest", "Located", "ERO Reprocessed Arrest",
  "Other efforts", "Other Task Force", "Worksite Enforcement", "Patrol Interior"
)

# Full administration years, split on the Jan 20 inauguration date. Defined once
# here so the same boundaries drive every table, including the national figures.
ADMIN_LEVELS <- c("Biden, late 2023 (partial)", "Biden's last year",
                  "Trump's first year", "Trump's second year (so far)")
assign_admin_period <- function(d) {
  factor(case_when(
    d >= as.Date("2026-01-20") ~ "Trump's second year (so far)",
    d >= as.Date("2025-01-20") ~ "Trump's first year",
    d >= as.Date("2024-01-20") ~ "Biden's last year",
    TRUE                       ~ "Biden, late 2023 (partial)"
  ), levels = ADMIN_LEVELS)
}

arrests_local <- arrests_local |>
  mutate(
    arrest_type = case_when(
      apprehension_method %in% CUSTODIAL_METHODS ~ "Custodial",
      apprehension_method %in% AT_LARGE_METHODS  ~ "Non-Custodial / At-Large",
      TRUE                                        ~ "Border / Entry"
    ),
    criminality_label = case_when(
      apprehension_criminality == "1 Convicted Criminal"         ~ "Convicted",
      apprehension_criminality == "2 Pending Criminal Charges"   ~ "Pending Charges",
      apprehension_criminality == "3 Other Immigration Violator" ~ "No Criminal Record",
      TRUE ~ "Unknown"
    ),
    admin_period = assign_admin_period(apprehension_date)
  )
```

---

## Data Check

Run this before trusting anything below. It surfaces the judgment calls baked
into your numbers so you can defend them — or catch a problem before it reaches
print.

```{r data-check}
# De-duplication across the three fiscal-year files
cat("Rows loaded:", n_raw_rows, "| after removing duplicates:", nrow(arrests_all),
    "(", n_raw_rows - nrow(arrests_all), "dropped )\n\n")

# How your filter mode pulled records
cat("Filter mode:", MY_FILTER_MODE, "selected", nrow(arrests_local), "arrests\n")
cat("  matched on state ==", MY_STATE, ":",
    sum(arrests_all$state == MY_STATE, na.rm = TRUE), "\n")
cat("  blank-state records added via AOR fallback:",
    sum(is.na(arrests_all$state) & arrests_all$toa_current_duty_aor == MY_AOR, na.rm = TRUE),
    "\n\n")

# Which states your AOR actually covers, from the authoritative ICE AOR county
# shapefile (reference/aor_states.csv) rather than inferred from the messy data.
aor_lookup <- read_csv(here("reference", "aor_states.csv"), show_col_types = FALSE)
my_key   <- str_squish(str_to_lower(str_remove_all(str_remove(MY_AOR, "Area of Responsibility"), "\\.")))
this_aor <- aor_lookup |> filter(aor_key == my_key)

if (nrow(this_aor) == 1) {
  covered <- str_to_upper(str_trim(unlist(str_split(this_aor$states, ";"))))
  cat(MY_AOR, "officially covers", this_aor$n_states,
      if (this_aor$n_states == 1) "state:" else "states:", this_aor$states, "\n")
  if (!str_to_upper(MY_STATE) %in% covered) {
    cat("  WARNING: MY_STATE (", MY_STATE, ") is not in this AOR — check your config pairing.\n")
  }
  if (this_aor$n_states > 1) {
    cat("  Multi-state AOR: state_only is safest; state_or_aor / state_and_aor can pull in",
        paste(setdiff(covered, str_to_upper(MY_STATE)), collapse = ", "), "\n")
  }
} else {
  cat("AOR not found in reference/aor_states.csv; skipping the geography check.\n")
}
cat("\n")

# Whether your state is split across several AORs. If so, a state-wide filter
# lumps separate metro regions together (e.g. New York State = NYC + Buffalo),
# which surprises reporters expecting just one city.
state_aors <- aor_lookup |>
  filter(map_lgl(states, ~ str_to_upper(MY_STATE) %in%
                   str_to_upper(str_trim(unlist(str_split(.x, ";")))))) |>
  pull(aor_name)
if (length(state_aors) > 1 && MY_FILTER_MODE != "state_and_aor") {
  cat("NOTE:", str_to_title(MY_STATE), "is split across", length(state_aors),
      "ICE AORs:", paste(state_aors, collapse = ", "), "\n")
  cat("  Mode", MY_FILTER_MODE, "includes all of them. To isolate one metro area",
      "(e.g. New York City, not Buffalo/upstate), set MY_FILTER_MODE = 'state_and_aor' and pick one AOR.\n\n")
}

# Apprehension methods that fall into the 'Border / Entry' catch-all
unmapped <- arrests_local |>
  filter(!apprehension_method %in% c(CUSTODIAL_METHODS, AT_LARGE_METHODS)) |>
  count(apprehension_method, sort = TRUE)
cat("Methods landing in 'Border / Entry' (", sum(unmapped$n), "rows ) — confirm they belong:\n")
if (nrow(unmapped) > 0) print(unmapped) else cat("  none\n")
cat("\n")

# Date coverage and partial-window flag
cat("Date range:", format(min(arrests_local$apprehension_date, na.rm = TRUE)),
    "to", format(max(arrests_local$apprehension_date, na.rm = TRUE)), "\n")
if (END_DATE > max(arrests_local$apprehension_date, na.rm = TRUE)) {
  cat("  END_DATE is past the data, so the most recent period is data-limited, not calendar-limited.\n")
}
n_future <- sum(arrests_all$apprehension_date > Sys.Date(), na.rm = TRUE)
if (n_future > 0) {
  cat("  ", n_future, "record(s) are dated in the future (data-entry errors); excluded by the END_DATE filter.\n")
}

# Small-base guard for the period breakdowns below
small <- bind_rows(
  arrests_local |> count(admin_period, arrest_type) |> rename(category = arrest_type),
  arrests_local |> count(admin_period, criminality_label) |> rename(category = criminality_label)
) |>
  filter(n > 0, n < 10)
if (nrow(small) > 0) {
  cat("\nSMALL-SAMPLE WARNING: these breakdown cells have fewer than 10 arrests.\n")
  cat("Don't lead with a percentage calculated off them:\n")
  print(small)
}
```

---

## At a Glance

A draft lede built automatically from the administration-year comparison below.
Treat it as a starting point to verify against the tables, not finished copy.

```{r at-a-glance, results='asis'}
# Full administration-year totals, each flagged complete or partial. The data
# starts in Oct 2023 and the current year is ongoing, so the first and last
# buckets are partial; year-over-year change is only meaningful between two
# complete years. Built once here and reused by the Summary table.
data_min <- min(arrests_local$apprehension_date, na.rm = TRUE)
data_max <- max(arrests_local$apprehension_date, na.rm = TRUE)
period_info <- tibble(
  admin_period = factor(ADMIN_LEVELS, levels = ADMIN_LEVELS),
  start = as.Date(c(NA, "2024-01-20", "2025-01-20", "2026-01-20")),
  end   = as.Date(c("2024-01-19", "2025-01-19", "2026-01-19", "2027-01-19"))
) |>
  mutate(complete = !is.na(start) & start >= data_min & end <= data_max)

admin_summary <- arrests_local |>
  count(admin_period, .drop = FALSE) |>
  left_join(period_info |> select(admin_period, complete), by = "admin_period") |>
  mutate(
    share    = n / sum(n),
    valid    = complete & lag(complete),
    vs_prior = if_else(valid, n - lag(n), NA_integer_),
    pct      = if_else(valid, (n - lag(n)) / lag(n), NA_real_)
  )

complete_years <- admin_summary |> filter(complete)
if (nrow(complete_years) >= 2) {
  latest <- complete_years[nrow(complete_years), ]
  prior  <- complete_years[nrow(complete_years) - 1, ]
  pctc   <- (latest$n - prior$n) / prior$n * 100
  dir    <- if (pctc >= 0) "rose" else "fell"

  nc <- arrests_local |> filter(criminality_label == "No Criminal Record") |>
    count(admin_period, .drop = FALSE)
  nc_val <- function(p) coalesce(nc$n[match(p, as.character(nc$admin_period))], 0L)
  top_type <- arrests_local |> filter(admin_period == latest$admin_period) |>
    count(arrest_type, sort = TRUE) |> slice(1)

  partial <- admin_summary |> filter(admin_period == "Trump's second year (so far)")
  partial_txt <- if (nrow(partial) == 1 && partial$n > 0)
    sprintf(" %s arrests have been logged so far in Trump's second year.", comma(partial$n)) else ""

  cat(sprintf(
    paste0("> **At a glance — %s.** ICE arrests in %s **%s %s%%** between %s (%s) and %s (%s). ",
           "%s arrests led the most recent full year (%s of %s), and arrests of people with no ",
           "criminal record went from %s to %s.%s\n"),
    MY_LABEL, MY_LABEL, dir, round(abs(pctc)),
    prior$admin_period, comma(prior$n), latest$admin_period, comma(latest$n),
    top_type$arrest_type, comma(top_type$n), comma(latest$n),
    comma(nc_val(as.character(prior$admin_period))), comma(nc_val(as.character(latest$admin_period))),
    partial_txt))
} else {
  cat("> **At a glance.** Not enough complete administration years in the data to draft a year-over-year lede.\n")
}
```

---

## Summary

Arrests by administration year (Jan 20 to Jan 19 of the following year). The
first and last periods are partial — the data starts in October 2023 and the
current year is ongoing — so the year-over-year change is shown only between the
two complete years.

```{r summary-table}
admin_summary |>
  transmute(
    `Administration period`  = admin_period,
    Arrests                  = comma(n),
    Share                    = percent(share, accuracy = 0.1),
    `Change vs. prior year`  = if_else(is.na(vs_prior), "—", comma(vs_prior)),
    `% change`               = if_else(is.na(pct), "—", percent(pct, accuracy = 0.1))
  ) |>
  knitr::kable(align = c("l", "r", "r", "r", "r"))
```

### Arrest Type by Administration Year

Each administration year as a column (the first and last are partial). This shows
whether the *mix* of arrests shifted, not just the total.

```{r type-by-period}
arrests_local |>
  count(admin_period, arrest_type, .drop = FALSE) |>
  filter(n > 0) |>
  pivot_wider(names_from = admin_period, values_from = n, values_fill = 0) |>
  rename(`Arrest Type` = arrest_type) |>
  knitr::kable()
```

### Criminality by Administration Year

```{r criminality-by-period}
arrests_local |>
  count(admin_period, criminality_label, .drop = FALSE) |>
  filter(n > 0) |>
  pivot_wider(names_from = admin_period, values_from = n, values_fill = 0) |>
  rename(`Criminality` = criminality_label) |>
  knitr::kable()
```

---

## How Does This Compare?

A local number means more against a baseline. This ranks your state by arrests
per 100,000 residents in the most recent complete administration year (live Census
ACS via tidycensus when a key is set, bundled 2024 estimates otherwise) and sets
your year-over-year change beside the national one.

```{r compare-national, results='asis'}
# Population for per-capita rates. With a free Census API key set (see the
# README, "Census API key"), this pulls live ACS estimates via tidycensus.
# Without a key, offline, or if the call fails, it falls back to the bundled
# 2024 estimates in reference/state_population.csv so the report still knits.
pop <- tryCatch({
  stopifnot(requireNamespace("tidycensus", quietly = TRUE),
            nzchar(Sys.getenv("CENSUS_API_KEY")))
  tidycensus::get_acs(geography = "state", variables = "B01003_001",
                      year = 2023, survey = "acs5") |>
    transmute(state = str_to_upper(NAME), population = estimate)
}, error = function(e) {
  read_csv(here("reference", "state_population.csv"), show_col_types = FALSE)
})

# The two most recent complete administration years drive the comparison.
cy <- admin_summary |> filter(complete)
latest_lab <- as.character(cy$admin_period[nrow(cy)])
prior_lab  <- as.character(cy$admin_period[nrow(cy) - 1])
loc_pct <- (cy$n[nrow(cy)] - cy$n[nrow(cy) - 1]) / cy$n[nrow(cy) - 1] * 100

# National totals on the same full administration-year basis.
natl    <- arrests_all |> mutate(admin_period = assign_admin_period(apprehension_date))
natl_by <- natl |> count(admin_period, .drop = FALSE)
nlv      <- function(p) natl_by$n[match(p, as.character(natl_by$admin_period))]
natl_pct <- (nlv(latest_lab) - nlv(prior_lab)) / nlv(prior_lab) * 100

# Per-capita ranking for the most recent complete year (a full 12-month rate).
per_state <- natl |>
  filter(admin_period == latest_lab, !is.na(state)) |>
  count(state) |>
  inner_join(pop, by = "state") |>
  mutate(per100k = n / population * 1e5) |>
  arrange(desc(per100k)) |>
  mutate(rank = row_number())
me <- per_state |> filter(state == MY_STATE)

if (nrow(cy) >= 2 && nrow(me) == 1) {
  cat(sprintf(
    paste0("> **How %s compares.** In %s (the most recent complete administration year), %s logged ",
           "**%.1f arrests per 100,000 residents** statewide, ranking **#%d of %d** states. %s's arrests ",
           "%s **%.0f%%** from %s, against a national change of %s%.0f%%.\n"),
    MY_LABEL, latest_lab, str_to_title(MY_STATE), me$per100k, me$rank, nrow(per_state),
    MY_LABEL, if (loc_pct >= 0) "rose" else "fell", abs(loc_pct), prior_lab,
    if (natl_pct >= 0) "up " else "down ", abs(natl_pct)))
  if (MY_FILTER_MODE == "state_and_aor") {
    cat(sprintf(paste0("\n*Per-capita rank is statewide (%s) since population is only available by state; ",
                       "your tables above cover only the %s within it. The year-over-year figure is for %s.*\n"),
                str_to_title(MY_STATE), MY_AOR, MY_LABEL))
  } else {
    cat("\n*Per-capita ranking naturally favors smaller and border states, so read it alongside the raw counts. All figures cover the broad DDP arrests universe, not custodial arrests only.*\n")
  }
} else {
  cat("> **How it compares.** Not enough complete years or no population match, so the comparison is unavailable.\n")
}
```

```{r compare-table}
per_state |>
  transmute(Rank = rank, State = str_to_title(state),
            `Per 100k` = round(per100k, 1), Arrests = comma(n)) |>
  filter(Rank <= 5 | State == str_to_title(MY_STATE)) |>
  knitr::kable()
```

---

## Monthly Trend

```{r monthly-trend}
monthly <- arrests_local |>
  mutate(month = floor_date(apprehension_date, "month")) |>
  count(month)

ggplot(monthly, aes(x = month, y = n)) +
  geom_line(linewidth = 1, color = "#2c6fad") +
  geom_point(size = 2, color = "#2c6fad") +
  geom_vline(
    xintercept = START_DATE,
    linetype = "dashed", color = "gray40"
  ) +
  annotate(
    "text", x = START_DATE,
    y = max(monthly$n) * 0.95,
    label = paste("Inauguration\n", format(START_DATE, "%b %d, %Y")),
    hjust = -0.05, size = 3, color = "gray40"
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = paste("ICE Arrests in", MY_LABEL, "by Month"),
    subtitle = "FY2024–FY2026",
    x = NULL, y = "Arrests",
    caption  = "Source: ICE via Deportation Data Project (FOIA)"
  ) +
  theme_minimal()
```

---

## Arrest Type Over Time

```{r type-trend}
monthly_type <- arrests_local |>
  mutate(month = floor_date(apprehension_date, "month")) |>
  count(month, arrest_type)

ggplot(monthly_type, aes(x = month, y = n, fill = arrest_type)) +
  geom_col() +
  geom_vline(xintercept = START_DATE, linetype = "dashed", color = "white") +
  scale_fill_manual(values = c(
    "Custodial"               = "#2c6fad",
    "Non-Custodial / At-Large"= "#c94040",
    "Border / Entry"          = "#888888"
  )) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = paste("ICE Arrests in", MY_LABEL, "by Type"),
    subtitle = "Dashed line = inauguration",
    x = NULL, y = "Arrests", fill = NULL,
    caption  = "Source: ICE via Deportation Data Project (FOIA)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## Top Facilities

```{r top-facilities}
arrests_local |>
  filter(
    !is.na(apprehension_site_landmark),
    !str_detect(apprehension_site_landmark,
                paste(ALL_EXCLUSIONS, collapse = "|"))
  ) |>
  count(apprehension_site_landmark, arrest_type, sort = TRUE) |>
  slice_head(n = 15) |>
  knitr::kable(col.names = c("Facility", "Type", "Arrests"))
```

---

## Detainers

```{r load-detainers}
load_detainers <- function(path) {
  read_xlsx(here(path), skip = 6) |>
    clean_names() |>
    mutate(
      across(where(is.character), ~ na_if(.x, "NA")),
      detainer_prepare_date = as_date(detainer_prepare_date)
    )
}

detainers_local <- bind_rows(
  load_detainers(DETAINERS_FY24),
  load_detainers(DETAINERS_FY25),
  load_detainers(DETAINERS_FY26)
) |>
  distinct() |>
  filter(facility_state == MY_STATE)

cat(MY_LABEL, "detainers:", nrow(detainers_local), "\n")
```

```{r detainer-compliance}
detainers_local |>
  group_by(detainer_facility) |>
  summarize(
    total    = n(),
    declined = sum(detainer_lift_reason == "Detainer Declined by LEA", na.rm = TRUE),
    decline_rate = percent(declined / total, accuracy = 0.1)
  ) |>
  arrange(desc(total)) |>
  slice_head(n = 15) |>
  knitr::kable(col.names = c("Facility", "Total Detainers", "Declined", "Decline Rate"))
```

---

## Methodology

Data comes from ICE arrest records obtained through a Freedom of Information Act lawsuit and compiled by the Deportation Data Project. It covers ICE arrests only — not Customs and Border Protection. `r MY_LABEL` figures reflect arrests in ICE's `r MY_AOR` filtered by state, across the full FY2024 to FY2026 range. Per-capita rates use Census Bureau 2024 estimates. Arrests from local jail cooperation programs (287(g) and CAP) are counted as custodial; community arrests are counted as non-custodial. After July 2025, ICE folded CAP program arrests into a single "Custodial Arrest" category, so pre- and post-July comparisons by method should be made carefully.

Arrests are grouped into administration years anchored on the January 20 inauguration anniversary, rather than calendar or fiscal years, so each period is directly comparable. The data begins in October 2023, so "Biden, late 2023" is a partial window, and the data ends partway through "Trump's second year," which is also partial. Raw totals for those two periods are not full-year figures. The "same time last year" comparison addresses this by trimming every administration year to the same number of elapsed days, measured from each January 20, so the most recent partial year is compared only against the matching window of earlier years.
