---
title: "ICE Arrests and Detainers — Local Analysis"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(readxl)
library(janitor)
library(scales)
library(here)        # resolves paths from the project root, however you run this

# here() anchors paths to the project root, so this works whether you knit or run chunks interactively
source(here("my_newsroom", "config.R"))
source(here("my_newsroom", "excluded_landmarks.R"))
```

---

## Load and Clean Data

```{r load-arrests}
load_arrests <- function(path) {
  read_xlsx(here(path), skip = 6) |>
    clean_names() |>
    mutate(
      across(where(is.character), ~ na_if(.x, "NA")),
      apprehension_date = as_date(apprehension_date)
    )
}

arrests_all <- bind_rows(
  load_arrests(ARRESTS_FY24),
  load_arrests(ARRESTS_FY25),
  load_arrests(ARRESTS_FY26)
)
n_raw_rows <- nrow(arrests_all)          # before de-duplication, for the Data Check
arrests_all <- distinct(arrests_all)
```

```{r filter-state}
# Filter modes: state_only (safest for multi-state AORs), state_or_aor (state plus blank-state AOR matches), state_and_aor (one AOR within a state, for sub-state breakouts like NYC)
arrests_local <- arrests_all |>
  filter(
    if (MY_FILTER_MODE == "state_only") {
      state == MY_STATE
    } else if (MY_FILTER_MODE == "state_and_aor") {
      state == MY_STATE & toa_current_duty_aor == MY_AOR
    } else {
      state == MY_STATE | (is.na(state) & toa_current_duty_aor == MY_AOR)
    }
  ) |>
  filter(apprehension_date <= END_DATE)   # full FY24-FY26 range; sliced into administration years below

cat(MY_LABEL, "arrests:", nrow(arrests_local), "\n")
```

```{r recode}
CUSTODIAL_METHODS <- c(
  "Custodial Arrest", "CAP Local Incarceration", "CAP Federal Incarceration",
  "CAP State Incarceration", "287(g) Program", "Probation and Parole",
  "Law Enforcement Agency Response Unit", "Other Agency (turned over to INS)",
  "Organized Crime Drug Enforcement Task Force", "Criminal Alien Program"
)

AT_LARGE_METHODS <- c(
  "Non-Custodial Arrest", "Located", "ERO Reprocessed Arrest",
  "Other efforts", "Other Task Force", "Worksite Enforcement", "Patrol Interior"
)

arrests_local <- arrests_local |>
  mutate(
    arrest_type = case_when(
      apprehension_method %in% CUSTODIAL_METHODS ~ "Custodial",
      apprehension_method %in% AT_LARGE_METHODS  ~ "Non-Custodial / At-Large",
      TRUE                                        ~ "Border / Entry"
    ),
    criminality_label = case_when(
      apprehension_criminality == "1 Convicted Criminal"         ~ "Convicted",
      apprehension_criminality == "2 Pending Criminal Charges"   ~ "Pending Charges",
      apprehension_criminality == "3 Other Immigration Violator" ~ "No Criminal Record",
      TRUE ~ "Unknown"
    ),
    # Administration years anchored on the Jan 20 inauguration anniversary; most-recent first
    admin_period = case_when(
      apprehension_date >= as.Date("2026-01-20") ~ "Trump's second year (so far)",
      apprehension_date >= as.Date("2025-01-20") ~ "Trump's first year",
      apprehension_date >= as.Date("2024-01-20") ~ "Biden's last year",
      TRUE                                       ~ "Biden, late 2023 (partial)"
    )
  ) |>
  mutate(admin_period = factor(admin_period, levels = c(
    "Biden, late 2023 (partial)",
    "Biden's last year",
    "Trump's first year",
    "Trump's second year (so far)"
  )))
```

---

## Data Check

Run this before trusting anything below. It surfaces the judgment calls baked
into your numbers so you can defend them — or catch a problem before it reaches
print.

```{r data-check}
# De-duplication across the three fiscal-year files
cat("Rows loaded:", n_raw_rows, "| after removing duplicates:", nrow(arrests_all),
    "(", n_raw_rows - nrow(arrests_all), "dropped )\n\n")

# How your filter mode pulled records
cat("Filter mode:", MY_FILTER_MODE, "selected", nrow(arrests_local), "arrests\n")
cat("  matched on state ==", MY_STATE, ":",
    sum(arrests_all$state == MY_STATE, na.rm = TRUE), "\n")
cat("  blank-state records added via AOR fallback:",
    sum(is.na(arrests_all$state) & arrests_all$toa_current_duty_aor == MY_AOR, na.rm = TRUE),
    "\n\n")

# Whether your AOR mixes in other states
aor_states <- arrests_all |>
  filter(toa_current_duty_aor == MY_AOR, !is.na(state)) |>
  count(state, sort = TRUE)
if (nrow(aor_states) > 1) {
  cat("NOTE:", MY_AOR, "spans", nrow(aor_states),
      "states. A state-anchored mode keeps your count clean; an AOR-only view would mix them.\n")
  print(aor_states)
  cat("\n")
}

# Apprehension methods that fall into the 'Border / Entry' catch-all
unmapped <- arrests_local |>
  filter(!apprehension_method %in% c(CUSTODIAL_METHODS, AT_LARGE_METHODS)) |>
  count(apprehension_method, sort = TRUE)
cat("Methods landing in 'Border / Entry' (", sum(unmapped$n), "rows ) — confirm they belong:\n")
if (nrow(unmapped) > 0) print(unmapped) else cat("  none\n")
cat("\n")

# Date coverage and partial-window flag
cat("Date range:", format(min(arrests_local$apprehension_date, na.rm = TRUE)),
    "to", format(max(arrests_local$apprehension_date, na.rm = TRUE)), "\n")
if (END_DATE > max(arrests_local$apprehension_date, na.rm = TRUE)) {
  cat("  END_DATE is past the data, so the most recent period is data-limited, not calendar-limited.\n")
}

# Small-base guard for the period breakdowns below
small <- bind_rows(
  arrests_local |> count(admin_period, arrest_type) |> rename(category = arrest_type),
  arrests_local |> count(admin_period, criminality_label) |> rename(category = criminality_label)
) |>
  filter(n > 0, n < 10)
if (nrow(small) > 0) {
  cat("\nSMALL-SAMPLE WARNING: these breakdown cells have fewer than 10 arrests.\n")
  cat("Don't lead with a percentage calculated off them:\n")
  print(small)
}
```

---

## Summary

Total arrests in each administration year. Note that the first and last
periods are partial windows (the data starts in October 2023 and ends
partway through Trump's second year), so compare the two full years to
each other, or use the "same time last year" table below.

```{r summary-table}
arrests_local |>
  count(admin_period, .drop = FALSE) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) |>
  knitr::kable(col.names = c("Administration period", "Arrests", "Share"))
```

### Same Time Last Year

The most recent administration year is only partial, so comparing its raw
total to a full prior year would understate it. To compare fairly, we line up
the **same calendar window** in every year: from each January 20 inauguration
anniversary through the latest date the data reaches (early March this year).

```{r yoy-window}
# How far into the current administration year does the data reach?
data_through  <- max(arrests_local$apprehension_date, na.rm = TRUE)
anniversaries <- as.Date(c("2024-01-20", "2025-01-20", "2026-01-20"))
current_anniv <- max(anniversaries[anniversaries <= data_through])
elapsed_days  <- as.integer(data_through - current_anniv)

# Flag arrests in the same elapsed-day window of their year; findInterval returns 0 for pre-2024 records, so map those to NA
arrests_same_window <- arrests_local |>
  mutate(
    anniv_idx    = findInterval(apprehension_date, anniversaries),
    period_start = if_else(anniv_idx == 0, as.Date(NA), anniversaries[pmax(anniv_idx, 1)])
  ) |>
  filter(!is.na(period_start),
         as.integer(apprehension_date - period_start) <= elapsed_days)

cat("Comparing Jan 20 to", format(data_through, "%b %d"),
    "in each administration year\n")

arrests_same_window |>
  count(admin_period, .drop = FALSE) |>
  filter(n > 0) |>
  mutate(
    vs_prior = n - lag(n),
    pct_change = percent((n - lag(n)) / lag(n), accuracy = 0.1)
  ) |>
  knitr::kable(col.names = c("Administration period", "Arrests (same window)",
                             "Change vs. prior year", "% change"))
```

### Arrest Type, Trump vs. Biden

Same elapsed-day window as above, broken out by administration so the periods
are directly comparable. This shows whether the *mix* of arrests shifted, not
just the total.

```{r type-by-period}
arrests_same_window |>
  count(admin_period, arrest_type, .drop = FALSE) |>
  filter(n > 0) |>
  pivot_wider(names_from = admin_period, values_from = n, values_fill = 0) |>
  rename(`Arrest Type` = arrest_type) |>
  knitr::kable()
```

### Criminality, Trump vs. Biden

```{r criminality-by-period}
arrests_same_window |>
  count(admin_period, criminality_label, .drop = FALSE) |>
  filter(n > 0) |>
  pivot_wider(names_from = admin_period, values_from = n, values_fill = 0) |>
  rename(`Criminality` = criminality_label) |>
  knitr::kable()
```

---

## Monthly Trend

```{r monthly-trend}
monthly <- arrests_local |>
  mutate(month = floor_date(apprehension_date, "month")) |>
  count(month)

ggplot(monthly, aes(x = month, y = n)) +
  geom_line(linewidth = 1, color = "#2c6fad") +
  geom_point(size = 2, color = "#2c6fad") +
  geom_vline(
    xintercept = START_DATE,
    linetype = "dashed", color = "gray40"
  ) +
  annotate(
    "text", x = START_DATE,
    y = max(monthly$n) * 0.95,
    label = paste("Inauguration\n", format(START_DATE, "%b %d, %Y")),
    hjust = -0.05, size = 3, color = "gray40"
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = paste("ICE Arrests in", MY_LABEL, "by Month"),
    subtitle = "FY2024–FY2026",
    x = NULL, y = "Arrests",
    caption  = "Source: ICE via Deportation Data Project (FOIA)"
  ) +
  theme_minimal()
```

---

## Arrest Type Over Time

```{r type-trend}
monthly_type <- arrests_local |>
  mutate(month = floor_date(apprehension_date, "month")) |>
  count(month, arrest_type)

ggplot(monthly_type, aes(x = month, y = n, fill = arrest_type)) +
  geom_col() +
  geom_vline(xintercept = START_DATE, linetype = "dashed", color = "white") +
  scale_fill_manual(values = c(
    "Custodial"               = "#2c6fad",
    "Non-Custodial / At-Large"= "#c94040",
    "Border / Entry"          = "#888888"
  )) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = paste("ICE Arrests in", MY_LABEL, "by Type"),
    subtitle = "Dashed line = inauguration",
    x = NULL, y = "Arrests", fill = NULL,
    caption  = "Source: ICE via Deportation Data Project (FOIA)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## Top Facilities

```{r top-facilities}
arrests_local |>
  filter(
    !is.na(apprehension_site_landmark),
    !str_detect(apprehension_site_landmark,
                paste(ALL_EXCLUSIONS, collapse = "|"))
  ) |>
  count(apprehension_site_landmark, arrest_type, sort = TRUE) |>
  slice_head(n = 15) |>
  knitr::kable(col.names = c("Facility", "Type", "Arrests"))
```

---

## Detainers

```{r load-detainers}
load_detainers <- function(path) {
  read_xlsx(here(path), skip = 6) |>
    clean_names() |>
    mutate(
      across(where(is.character), ~ na_if(.x, "NA")),
      detainer_prepare_date = as_date(detainer_prepare_date)
    )
}

detainers_local <- bind_rows(
  load_detainers(DETAINERS_FY24),
  load_detainers(DETAINERS_FY25),
  load_detainers(DETAINERS_FY26)
) |>
  distinct() |>
  filter(facility_state == MY_STATE)

cat(MY_LABEL, "detainers:", nrow(detainers_local), "\n")
```

```{r detainer-compliance}
detainers_local |>
  group_by(detainer_facility) |>
  summarize(
    total    = n(),
    declined = sum(detainer_lift_reason == "Detainer Declined by LEA", na.rm = TRUE),
    decline_rate = percent(declined / total, accuracy = 0.1)
  ) |>
  arrange(desc(total)) |>
  slice_head(n = 15) |>
  knitr::kable(col.names = c("Facility", "Total Detainers", "Declined", "Decline Rate"))
```

---

## Methodology

Data comes from ICE arrest records obtained through a Freedom of Information Act lawsuit and compiled by the Deportation Data Project. It covers ICE arrests only — not Customs and Border Protection. `r MY_LABEL` figures reflect arrests in ICE's `r MY_AOR` filtered by state, across the full FY2024 to FY2026 range. Per-capita rates use Census Bureau 2024 estimates. Arrests from local jail cooperation programs (287(g) and CAP) are counted as custodial; community arrests are counted as non-custodial. After July 2025, ICE folded CAP program arrests into a single "Custodial Arrest" category, so pre- and post-July comparisons by method should be made carefully.

Arrests are grouped into administration years anchored on the January 20 inauguration anniversary, rather than calendar or fiscal years, so each period is directly comparable. The data begins in October 2023, so "Biden, late 2023" is a partial window, and the data ends partway through "Trump's second year," which is also partial. Raw totals for those two periods are not full-year figures. The "same time last year" comparison addresses this by trimming every administration year to the same number of elapsed days, measured from each January 20, so the most recent partial year is compared only against the matching window of earlier years.
