---
title: "ICE Arrests and Detainers — Local Analysis"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(readxl)
library(janitor)
library(scales)
library(here)        # resolves paths from the project root, however you run this

# Load your configuration and exclusions. here() anchors to the project
# root (the folder with the .Rproj file), so these work whether you knit
# the file or run the chunks interactively in RStudio.
source(here("my_newsroom", "config.R"))
source(here("my_newsroom", "excluded_landmarks.R"))
```

---

## Load and Clean Data

```{r load-arrests}
load_arrests <- function(path) {
  read_xlsx(here(path), skip = 6) |>
    clean_names() |>
    mutate(
      across(where(is.character), ~ na_if(.x, "NA")),
      apprehension_date = as_date(apprehension_date)
    )
}

arrests_all <- bind_rows(
  load_arrests(ARRESTS_FY24),
  load_arrests(ARRESTS_FY25),
  load_arrests(ARRESTS_FY26)
) |>
  distinct()
```

```{r filter-state}
# Primary filter: state field
# Fallback: AOR field (catches cases where state is blank but AOR is correct)
arrests_local <- arrests_all |>
  filter(
    state == MY_STATE |
    (is.na(state) & toa_current_duty_aor == MY_AOR)
  ) |>
  # Keep the full FY24–FY26 range. We slice it into administration years
  # below; the lower bound is the start of the data, not a fixed date.
  filter(apprehension_date <= END_DATE)

cat(MY_LABEL, "arrests:", nrow(arrests_local), "\n")
```

```{r recode}
CUSTODIAL_METHODS <- c(
  "Custodial Arrest", "CAP Local Incarceration", "CAP Federal Incarceration",
  "CAP State Incarceration", "287(g) Program", "Probation and Parole",
  "Law Enforcement Agency Response Unit", "Other Agency (turned over to INS)",
  "Organized Crime Drug Enforcement Task Force", "Criminal Alien Program"
)

AT_LARGE_METHODS <- c(
  "Non-Custodial Arrest", "Located", "ERO Reprocessed Arrest",
  "Other efforts", "Other Task Force", "Worksite Enforcement", "Patrol Interior"
)

arrests_local <- arrests_local |>
  mutate(
    arrest_type = case_when(
      apprehension_method %in% CUSTODIAL_METHODS ~ "Custodial",
      apprehension_method %in% AT_LARGE_METHODS  ~ "Non-Custodial / At-Large",
      TRUE                                        ~ "Border / Entry"
    ),
    criminality_label = case_when(
      apprehension_criminality == "1 Convicted Criminal"         ~ "Convicted",
      apprehension_criminality == "2 Pending Criminal Charges"   ~ "Pending Charges",
      apprehension_criminality == "3 Other Immigration Violator" ~ "No Criminal Record",
      TRUE ~ "Unknown"
    ),
    # Administration years, anchored on the Jan 20 inauguration anniversary
    # so each period is directly comparable. Most-recent first in case_when.
    admin_period = case_when(
      apprehension_date >= as.Date("2026-01-20") ~ "Trump's second year (so far)",
      apprehension_date >= as.Date("2025-01-20") ~ "Trump's first year",
      apprehension_date >= as.Date("2024-01-20") ~ "Biden's last year",
      TRUE                                       ~ "Biden, late 2023 (partial)"
    )
  ) |>
  mutate(admin_period = factor(admin_period, levels = c(
    "Biden, late 2023 (partial)",
    "Biden's last year",
    "Trump's first year",
    "Trump's second year (so far)"
  )))
```

---

## Summary

Total arrests in each administration year. Note that the first and last
periods are partial windows (the data starts in October 2023 and ends
partway through Trump's second year), so compare the two full years to
each other, or use the "same time last year" table below.

```{r summary-table}
arrests_local |>
  count(admin_period, .drop = FALSE) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) |>
  knitr::kable(col.names = c("Administration period", "Arrests", "Share"))
```

### Same Time Last Year

The most recent administration year is only partial, so comparing its raw
total to a full prior year would understate it. To compare fairly, we trim
every administration year to the **same number of days** that have elapsed
in the current one, measured from each January 20 anniversary.

```{r yoy-window}
# How far into the current administration year does the data reach?
data_through  <- max(arrests_local$apprehension_date, na.rm = TRUE)
anniversaries <- as.Date(c("2024-01-20", "2025-01-20", "2026-01-20"))
current_anniv <- max(anniversaries[anniversaries <= data_through])
elapsed_days  <- as.integer(data_through - current_anniv)

# Flag arrests that fall in the same elapsed-day window of their own year.
# findInterval returns 0 for late-2023 records (before the first
# anniversary); map those to NA so they drop out cleanly.
arrests_same_window <- arrests_local |>
  mutate(
    anniv_idx    = findInterval(apprehension_date, anniversaries),
    period_start = if_else(anniv_idx == 0, as.Date(NA), anniversaries[pmax(anniv_idx, 1)])
  ) |>
  filter(!is.na(period_start),
         as.integer(apprehension_date - period_start) <= elapsed_days)

cat("Comparing the first", elapsed_days, "days of each administration year",
    "(through", format(data_through, "%b %d"), ")\n")

arrests_same_window |>
  count(admin_period, .drop = FALSE) |>
  filter(n > 0) |>
  mutate(
    vs_prior = n - lag(n),
    pct_change = percent((n - lag(n)) / lag(n), accuracy = 0.1)
  ) |>
  knitr::kable(col.names = c("Administration period", "Arrests (same window)",
                             "Change vs. prior year", "% change"))
```

```{r type-table}
arrests_local |>
  count(arrest_type, sort = TRUE) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) |>
  knitr::kable(col.names = c("Arrest Type", "Count", "Share"))
```

```{r criminality-table}
arrests_local |>
  count(criminality_label, sort = TRUE) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) |>
  knitr::kable(col.names = c("Criminality", "Count", "Share"))
```

---

## Monthly Trend

```{r monthly-trend}
monthly <- arrests_local |>
  mutate(month = floor_date(apprehension_date, "month")) |>
  count(month)

ggplot(monthly, aes(x = month, y = n)) +
  geom_line(linewidth = 1, color = "#2c6fad") +
  geom_point(size = 2, color = "#2c6fad") +
  geom_vline(
    xintercept = START_DATE,
    linetype = "dashed", color = "gray40"
  ) +
  annotate(
    "text", x = START_DATE,
    y = max(monthly$n) * 0.95,
    label = paste("Inauguration\n", format(START_DATE, "%b %d, %Y")),
    hjust = -0.05, size = 3, color = "gray40"
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = paste("ICE Arrests in", MY_LABEL, "by Month"),
    subtitle = "FY2024–FY2026",
    x = NULL, y = "Arrests",
    caption  = "Source: DHS via Deportation Data Project (FOIA)"
  ) +
  theme_minimal()
```

---

## Arrest Type Over Time

```{r type-trend}
monthly_type <- arrests_local |>
  mutate(month = floor_date(apprehension_date, "month")) |>
  count(month, arrest_type)

ggplot(monthly_type, aes(x = month, y = n, fill = arrest_type)) +
  geom_col() +
  geom_vline(xintercept = START_DATE, linetype = "dashed", color = "white") +
  scale_fill_manual(values = c(
    "Custodial"               = "#2c6fad",
    "Non-Custodial / At-Large"= "#c94040",
    "Border / Entry"          = "#888888"
  )) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = paste("ICE Arrests in", MY_LABEL, "by Type"),
    subtitle = "Dashed line = inauguration",
    x = NULL, y = "Arrests", fill = NULL,
    caption  = "Source: DHS via Deportation Data Project (FOIA)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## Top Facilities

```{r top-facilities}
arrests_local |>
  filter(
    !is.na(apprehension_site_landmark),
    !str_detect(apprehension_site_landmark,
                paste(ALL_EXCLUSIONS, collapse = "|"))
  ) |>
  count(apprehension_site_landmark, arrest_type, sort = TRUE) |>
  slice_head(n = 15) |>
  knitr::kable(col.names = c("Facility", "Type", "Arrests"))
```

---

## Detainers

```{r load-detainers}
load_detainers <- function(path) {
  read_xlsx(here(path), skip = 6) |>
    clean_names() |>
    mutate(
      across(where(is.character), ~ na_if(.x, "NA")),
      detainer_prepare_date = as_date(detainer_prepare_date)
    )
}

detainers_local <- bind_rows(
  load_detainers(DETAINERS_FY24),
  load_detainers(DETAINERS_FY25),
  load_detainers(DETAINERS_FY26)
) |>
  distinct() |>
  filter(facility_state == MY_STATE)

cat(MY_LABEL, "detainers:", nrow(detainers_local), "\n")
```

```{r detainer-compliance}
detainers_local |>
  group_by(detainer_facility) |>
  summarize(
    total    = n(),
    declined = sum(detainer_lift_reason == "Detainer Declined by LEA", na.rm = TRUE),
    decline_rate = percent(declined / total, accuracy = 0.1)
  ) |>
  arrange(desc(total)) |>
  slice_head(n = 15) |>
  knitr::kable(col.names = c("Facility", "Total Detainers", "Declined", "Decline Rate"))
```

---

## Methodology

Data comes from ICE arrest records obtained through a Freedom of Information Act lawsuit and compiled by the Deportation Data Project. It covers ICE arrests only — not Customs and Border Protection. `r MY_LABEL` figures reflect arrests in ICE's `r MY_AOR` filtered by state, across the full FY2024 to FY2026 range. Per-capita rates use Census Bureau 2024 estimates. Arrests from local jail cooperation programs (287(g) and CAP) are counted as custodial; community arrests are counted as non-custodial. After July 2025, ICE folded CAP program arrests into a single "Custodial Arrest" category, so pre- and post-July comparisons by method should be made carefully.

Arrests are grouped into administration years anchored on the January 20 inauguration anniversary, rather than calendar or fiscal years, so each period is directly comparable. The data begins in October 2023, so "Biden, late 2023" is a partial window, and the data ends partway through "Trump's second year," which is also partial. Raw totals for those two periods are not full-year figures. The "same time last year" comparison addresses this by trimming every administration year to the same number of elapsed days, measured from each January 20, so the most recent partial year is compared only against the matching window of earlier years.
