---
title: "ICE Arrests and Detainers — Local Analysis"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(readxl)
library(janitor)
library(scales)

# Load your configuration and exclusions.
# Paths are relative to the project root because _quarto.yml sets
# execute-dir: project (so data/raw/ paths in config.R resolve too).
source("my_newsroom/config.R")
source("my_newsroom/excluded_landmarks.R")
```

---

## Load and Clean Data

```{r load-arrests}
load_arrests <- function(path) {
  read_xlsx(path, skip = 6) |>
    clean_names() |>
    mutate(
      across(where(is.character), ~ na_if(.x, "NA")),
      apprehension_date = as_date(apprehension_date)
    )
}

arrests_all <- bind_rows(
  load_arrests(ARRESTS_FY24),
  load_arrests(ARRESTS_FY25),
  load_arrests(ARRESTS_FY26)
) |>
  distinct()
```

```{r filter-state}
# Primary filter: state field
# Fallback: AOR field (catches cases where state is blank but AOR is correct)
arrests_local <- arrests_all |>
  filter(
    state == MY_STATE |
    (is.na(state) & toa_current_duty_aor == MY_AOR)
  ) |>
  filter(
    apprehension_date >= as.Date("2024-10-01"),
    apprehension_date <= END_DATE
  )

cat(MY_LABEL, "arrests:", nrow(arrests_local), "\n")
```

```{r recode}
CUSTODIAL_METHODS <- c(
  "Custodial Arrest", "CAP Local Incarceration", "CAP Federal Incarceration",
  "CAP State Incarceration", "287(g) Program", "Probation and Parole",
  "Law Enforcement Agency Response Unit", "Other Agency (turned over to INS)",
  "Organized Crime Drug Enforcement Task Force", "Criminal Alien Program"
)

AT_LARGE_METHODS <- c(
  "Non-Custodial Arrest", "Located", "ERO Reprocessed Arrest",
  "Other efforts", "Other Task Force", "Worksite Enforcement", "Patrol Interior"
)

arrests_local <- arrests_local |>
  mutate(
    arrest_type = case_when(
      apprehension_method %in% CUSTODIAL_METHODS ~ "Custodial",
      apprehension_method %in% AT_LARGE_METHODS  ~ "Non-Custodial / At-Large",
      TRUE                                        ~ "Border / Entry"
    ),
    criminality_label = case_when(
      apprehension_criminality == "1 Convicted Criminal"         ~ "Convicted",
      apprehension_criminality == "2 Pending Criminal Charges"   ~ "Pending Charges",
      apprehension_criminality == "3 Other Immigration Violator" ~ "No Criminal Record",
      TRUE ~ "Unknown"
    ),
    era = if_else(
      apprehension_date >= START_DATE,
      paste("Post-inauguration\n(from", format(START_DATE, "%b %d, %Y"), ")"),
      paste("Pre-inauguration\n(before", format(START_DATE, "%b %d, %Y"), ")")
    )
  )
```

---

## Summary

```{r summary-table}
arrests_local |>
  mutate(year = year(apprehension_date)) |>
  count(year) |>
  knitr::kable(col.names = c("Year", "Arrests"))
```

```{r type-table}
arrests_local |>
  count(arrest_type, sort = TRUE) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) |>
  knitr::kable(col.names = c("Arrest Type", "Count", "Share"))
```

```{r criminality-table}
arrests_local |>
  count(criminality_label, sort = TRUE) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) |>
  knitr::kable(col.names = c("Criminality", "Count", "Share"))
```

---

## Monthly Trend

```{r monthly-trend}
monthly <- arrests_local |>
  mutate(month = floor_date(apprehension_date, "month")) |>
  count(month)

ggplot(monthly, aes(x = month, y = n)) +
  geom_line(linewidth = 1, color = "#2c6fad") +
  geom_point(size = 2, color = "#2c6fad") +
  geom_vline(
    xintercept = START_DATE,
    linetype = "dashed", color = "gray40"
  ) +
  annotate(
    "text", x = START_DATE,
    y = max(monthly$n) * 0.95,
    label = paste("Inauguration\n", format(START_DATE, "%b %d, %Y")),
    hjust = -0.05, size = 3, color = "gray40"
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = paste("ICE Arrests in", MY_LABEL, "by Month"),
    subtitle = "FY2024–FY2026",
    x = NULL, y = "Arrests",
    caption  = "Source: DHS via Deportation Data Project (FOIA)"
  ) +
  theme_minimal()
```

---

## Arrest Type Over Time

```{r type-trend}
monthly_type <- arrests_local |>
  mutate(month = floor_date(apprehension_date, "month")) |>
  count(month, arrest_type)

ggplot(monthly_type, aes(x = month, y = n, fill = arrest_type)) +
  geom_col() +
  geom_vline(xintercept = START_DATE, linetype = "dashed", color = "white") +
  scale_fill_manual(values = c(
    "Custodial"               = "#2c6fad",
    "Non-Custodial / At-Large"= "#c94040",
    "Border / Entry"          = "#888888"
  )) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = paste("ICE Arrests in", MY_LABEL, "by Type"),
    subtitle = "Dashed line = inauguration",
    x = NULL, y = "Arrests", fill = NULL,
    caption  = "Source: DHS via Deportation Data Project (FOIA)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## Top Facilities

```{r top-facilities}
arrests_local |>
  filter(
    !is.na(apprehension_site_landmark),
    !str_detect(apprehension_site_landmark,
                paste(ALL_EXCLUSIONS, collapse = "|"))
  ) |>
  count(apprehension_site_landmark, arrest_type, sort = TRUE) |>
  slice_head(n = 15) |>
  knitr::kable(col.names = c("Facility", "Type", "Arrests"))
```

---

## Detainers

```{r load-detainers}
load_detainers <- function(path) {
  read_xlsx(path, skip = 6) |>
    clean_names() |>
    mutate(
      across(where(is.character), ~ na_if(.x, "NA")),
      detainer_prepare_date = as_date(detainer_prepare_date)
    )
}

detainers_local <- bind_rows(
  load_detainers(DETAINERS_FY24),
  load_detainers(DETAINERS_FY25),
  load_detainers(DETAINERS_FY26)
) |>
  distinct() |>
  filter(facility_state == MY_STATE)

cat(MY_LABEL, "detainers:", nrow(detainers_local), "\n")
```

```{r detainer-compliance}
detainers_local |>
  group_by(detainer_facility) |>
  summarize(
    total    = n(),
    declined = sum(detainer_lift_reason == "Detainer Declined by LEA", na.rm = TRUE),
    decline_rate = percent(declined / total, accuracy = 0.1)
  ) |>
  arrange(desc(total)) |>
  slice_head(n = 15) |>
  knitr::kable(col.names = c("Facility", "Total Detainers", "Declined", "Decline Rate"))
```

---

## Methodology

Data comes from ICE arrest records obtained through a Freedom of Information Act lawsuit and compiled by the Deportation Data Project. It covers ICE arrests only — not Customs and Border Protection. `r MY_LABEL` figures reflect arrests in ICE's `r MY_AOR` filtered by state. Per-capita rates use Census Bureau 2024 estimates. Arrests from local jail cooperation programs (287(g) and CAP) are counted as custodial; community arrests are counted as non-custodial. After July 2025, ICE folded CAP program arrests into a single "Custodial Arrest" category, so pre- and post-July comparisons by method should be made carefully.
